---
title: "Assignment 04"
author: "Nick Stabile"
date: "9/22/2020"
output: html_document
---
Rmarkdown
Create and reasonably format an RMarkdown document for this analysis. Specifically:
• Include a title and your name as the author within the opening YAML;
• Use a hyperlink to link to the source of your data;
• Use appropriate headers to signify each visualization;
• Include code chunks for data manipulation and visualization so we can understand the code and analysis
you ran to create the graphs;
• Hide warning messages and unnecessary printing of data;

Four ggplot2 graphs
Your analysis should have four data visualizations of distinct graph types, made with ggplot2. Across all four graphs, use a total of:
• Six different aesthetics;
• Six different non-aesthetic options;
• Five different geoms;
• Two different scales (meaning change the default scale used for at least two aesthetics).

Further, each graph must include:
• Correct usage of all visual encodings;
• Appropriate data sourcing;
• Proper labeling of ALL visual encodings;
• An appropriate title and subtitle;
• The co

Next steps:
- Read in file
- figure out how to sum the projects with the same address to create total project cost variable (like I did in Stata)
- maybe think about adding in the market rate data from here: https://octo.quickbase.com/db/bit4krbdh?a=q&qid=44
  - also email DMPED about it
- Create ratios of incentive amount to affordable housing production
  - Basically affordable units per dollar spent on the incentive
  - When I have the total project amounts I can divide the particular incentive amount by the total     project amount (for projects with only 1 funding source this will just be 1) and then divide by the affordable units 
- Look at how many projects had multiple funding sources
- Look at which funding sources are most often used in combination (are there certain ones used more often alone or in combination?)
- Look at projects that didn't have any affordable housing (maybe manually)
- Re-read the report to see what other things they did and what I can build on

Graph 1: geom_col, investment by year by incentive name
Graph 2: investment dollars per affordable units (total) by incentive name
Graph 3: geom_dotplot deeply affordable units produced by incentive name (maybe use facets?)
Graph 4: geom_hist, number of incentives used for projects


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(haven)
library(scales)
```

```{r}

dc_housing_investment <- read_csv("dc-housing-investment.csv") %>% 
  rename(
    year = `FY`,
    incentive_name = `Incentive Name`,
    recipient_name = `Recipient Name`,
    incentive_amount = `Incentive Amount`, 
    investment_address = `MAR Address of Investment`,
    ward = `MAR_Ward`,
    ami_30 = `30% AMI`,
    ami_50 = `50% AMI`,
    ami_60 = `60% AMI`,
    ami_80 = `80% AMI`,
    total_affordable = `Total Affordable Units Produced or Preserved`,
    ) %>% 
  mutate(
    investment_address = str_to_upper(investment_address),
    ami_30 = as.numeric(ami_30),
    ami_50 = as.numeric(ami_50),
    ami_60 = as.numeric(ami_60),
    ami_80 = as.numeric(ami_80),
    total_affordable = as.numeric(total_affordable),
    incentive_name=replace(incentive_name, incentive_name=="HFA Revenue Bond Issuance", "HFA Revenue Bond")
  )

total_proj_data <- dc_housing_investment %>% 
  group_by(investment_address) %>% 
  summarize(total_proj_investment = sum(incentive_amount))

dc_housing_investment <- 
  left_join(dc_housing_investment, total_proj_data, by = "investment_address")

dc_housing_investment <- dc_housing_investment %>% 
  mutate(
    investment_ratio = incentive_amount/total_proj_investment,
    total_affordable_ratio = investment_ratio*total_affordable,
    ami_30_ratio = investment_ratio*ami_30,
    ami_50_ratio = investment_ratio*ami_50,
    ami_60_ratio = investment_ratio*ami_60,
    ami_80_ratio = investment_ratio*ami_80,
    total_affordable_efficiency = if_else(total_affordable_ratio > 0, incentive_amount/total_affordable_ratio, 0),
    ami_30_efficiency = if_else(ami_30_ratio > 0, incentive_amount/ami_30_ratio, 0),
    ami_50_efficiency = if_else(ami_50_ratio > 0, incentive_amount/ami_50_ratio, 0),
    ami_60_efficiency = if_else(ami_60_ratio > 0, incentive_amount/ami_60_ratio, 0),
    ami_80_efficiency = if_else(ami_80_ratio > 0, incentive_amount/ami_80_ratio, 0),
    )

dc_housing_investment %>% 
  group_by(incentive_name) %>% 
  summarize(incentive_total_affordable = sum(total_affordable_ratio, na.rm = TRUE)) %>%
  filter(incentive_total_affordable > 100) %>% 
  ggplot() +
  geom_col(mapping = aes(
    x = reorder(str_wrap(incentive_name, 20), -incentive_total_affordable), 
    y = incentive_total_affordable)) +
  labs(
    x = "Investment Program",
    y = "Affordable Housing Units"
  ) + 
  coord_flip()

dc_housing_investment %>% 
  group_by(incentive_name) %>% 
  summarize(incentive_avg_efficiency = mean(total_affordable_efficiency, na.rm = TRUE)) %>%
  filter(incentive_avg_efficiency > 0) %>% 
  ggplot() +
  geom_col(mapping = aes(
    x = reorder(str_wrap(incentive_name, 20), -incentive_avg_efficiency), 
    y = incentive_avg_efficiency)) +
  scale_y_continuous(labels = comma) +
  labs(
    x = "Investment Program",
    y = "Investment Dollars Per Unit"
  ) + 
  coord_flip()

```
